---
title: "Getting started with `mcbarycenter` package"
author: James Peng, Florian Stijven, Linbo Wang, Peter Gilbert
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(mcbarycenter)
```

# Introduction

The `mcbarycenter` package implements a method for estimating the quantile
function of the Wasserstein barycenter using the marginal-constructed
barycenter (MCB) procedure. Briefly, the estimator can be used in sparse
sampling settings, in which each distribution is observed only through a
potentially small i.i.d. sample. The estimator works by first estimating a
series of binomial mixing distributions using a beta parametric specification,
a spline parametric specification, or the NPMLE. It then combines those mixing
distributions to estimate the Wasserstein barycenter.

For more details, please see [cite paper].

The `mcbarycenter` package estimates the quantile function of the Wasserstein
barycenter using either the empirical barycenter or the MCB estimator. In
addition, the package provides tools for calculating quantile differences,
graphing results, and carrying out hypothesis tests to compare barycenters
across groups of distributions.

Some functionality depends on external packages. In particular, `REBayes` and
`Rmosek` can be used to make the NPMLE workflow run faster. To implement the spline
mixing distribution estimator, we used code from  `deconvolveR`, but rewrote the
relevant function to make it faster for this package's use case. 

The examples below use the bundled `sparse_dist_data` data set.

```{r}
data("sparse_dist_data")
head(sparse_dist_data)
```

Note that the data set must be in long format, with one column specifying the
ids of the distributions and another column specifying the i.i.d. samples from
those distributions.

# Estimating the Wasserstein barycenter quantile function

## Empirical Wasserstein barycenter

The `empbary()` function computes empirical quantiles for each subject and then
averages those quantiles pointwise within each group. 

Arguments:

- `df`: the input data frame.
- `id_col`: the column identifying subjects or groups.
- `val_col`: the column containing the numeric observations.
- `weight_col`: an optional column containing group weights; if omitted, the
  empirical barycenter is unweighted.
- `alpha_grid`: the grid of quantile levels to estimate.
- `quantile_type`: the `type` argument passed to `stats::quantile()`.

Returns: 

- An `empbary_result` object containing the estimated quantile curve in
  `res`, an estimated covariance matrix in `cov`, and the standardized input
  data in `data`.
- This objected can be plotted with the function `graph_quantiles()`. 

Example:

```{r}
sample_grp0 <- subset(sparse_dist_data, group == 0)
sample_grp1 <- subset(sparse_dist_data, group == 1)

emp_res_grp0 <- empbary(
  df = sample_grp0,
  id_col = "id",
  val_col = "value"
)

emp_res_grp1 <- empbary(
  df = sample_grp1,
  id_col = "id",
  val_col = "value"
)

head(emp_res_grp0$res)
head(emp_res_grp1$res)
```


## Marginal-constructed barycenter

The `mcbary()` function estimates the barycenter quantile function using the
marginal-corrected barycenter procedure.

Arguments:

- `df`, `id_val`, `val_col`: same arguments as in `empbary()` function. 
- `method`: the mixture estimation method; options include `"spline"`,
  `"npmle"`, `"raw"`, and `"beta"`.
- `x_grid`: the threshold grid to use when `cutpoints` is not supplied.
- `cutpoints`: the number of evenly spaced grid points constructed from the
  observed data range; either `cutpoints` or `x_grid` must be supplied.
- `bootstrap_samples`: the number of bootstrap resamples used for uncertainty
  estimation.
- `alpha_grid`: the grid of quantile levels to estimate.
- `use_midpoint`: whether to use interval midpoints when constructing the
  quantile distribution.
- `estimate_endpoints`: whether to estimate the endpoint support regions.
- `use_isotonic_dist`: whether to enforce monotonicity on the estimated distribution 
of quantiles before converting them to quantile distributions.
- `use_isotonic_output`: whether to enforce monotonicity on the final
  estimated quantile curve and confidence bounds.
- `weight_col`: an optional column containing id-level weights.
- `progress`: whether to display a bootstrap progress bar.
- `ci_level`: the confidence level used for reported intervals.
- `...`: additional arguments passed to the mixture estimation routine.

Returns: 

- An `mcbary_result` object containing the estimated quantile curve and
  uncertainty summaries in `res`, the bootstrap covariance matrix in `cov`,
  the estimated mixing distributions in `mixtures`, the chosen estimation
  method in `method`, and the standardized input data in `data`.

```{r}
mcb_res_grp0 <- mcbary(
  df = sample_grp0,
  id_col = "id",
  val_col = "value",
  method = "beta",
  cutpoints = 100,
  use_midpoint = TRUE,
  estimate_endpoints = TRUE,
  bootstrap_samples = 20,
  progress = FALSE
)

mcb_res_grp1 <- mcbary(
  df = sample_grp1,
  id_col = "id",
  val_col = "value",
  method = "beta",
  cutpoints = 100,
  use_midpoint = TRUE,
  estimate_endpoints = TRUE,
  bootstrap_samples = 20,
  progress = FALSE
)

head(mcb_res_grp0$res)
head(mcb_res_grp1$res)
```

# Graphing tools

The package includes helpers for plotting both the estimated barycenter
quantiles and the intermediate mixing distributions.

The graph below displays the empirical barycenters and the MCB estimators 
for groups 0 and 1. 

```{r}
graph_quantiles(
  emp_grp0 = emp_res_grp0,
  emp_grp1 = emp_res_grp1,
  mcb_grp0 = mcb_res_grp0,
  mcb_grp1 = mcb_res_grp1
)
```

```{r}
graph_mixtures(mcb_res_grp0)
```

# Comparing Wasserstein barycenters

To compare barycenters between groups of observations, we can use the `quantile_diff()`
function, which calculates the quantile differences between the two groups.

For example, we can compute the difference between the estimated MCB quantile
functions for groups 0 and 1, and then graph the resulting quantile
difference curve.

```{r}
quantile_diff_res <- quantile_diff(mcb_res_grp0, mcb_res_grp1)

head(quantile_diff_res)
```

```{r}
graph_quantiles(mcb_difference = quantile_diff_res)
```

To compare barycenters across samples, we first estimate each barycenter on the
same quantile grid and then apply `equalbary_test()`. Here we compare the two
groups in `sparse_dist_data`. 

```{r}
comparison <- equalbary_test(mcb_res_grp0, mcb_res_grp1, B = 1000, seed = 1)

comparison$sup_norm$pval
comparison$int_sq_norm$pval
comparison$hotelling$pval
```
