---
title: "Getting started with `mcbarycenter` package"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(mcbarycenter)
```

# Introduction

Placeholder description of the method. This section will be filled in with a
more complete overview later.

`mcbarycenter` implements the method for calculating the quantile function of
barycenters using both the empirical barycenter and the MCB estimator. The
package also provides tools for calculating quantile differences, graphing
results, and carrying out hypothesis tests to compare barycenters across groups 
of distributions.

Some functionality depends on external packages. In particular, `REBayes` and
`Rmosek` can be used to make the NPMLE workflow run faster. We also use ideas
from `deconvolveR`, but rewrote the relevant function to make it faster for the
setting used in this package.

The examples below use the bundled `sparse_dist_data` data set.

```{r}
data("sparse_dist_data")
head(sparse_dist_data)
```

# Estimating the barycenter quantile function

## Empirical Barycenter

The `empbary()` function computes empirical quantiles within each subject or
group and then averages those quantiles across groups.

Arguments:

- `df`: the input data frame.
- `id_col`: the column identifying subjects or groups.
- `val_col`: the column containing the numeric observations.
- `weight_col`: an optional column containing group weights; if omitted, the
  empirical barycenter is unweighted.
- `alpha_grid`: the grid of quantile levels to estimate.
- `quantile_type`: the `type` argument passed to `stats::quantile()`.

Example:

```{r}
emp_res <- empbary(
  df = sparse_dist_data,
  id_col = "id",
  val_col = "value"
)

head(emp_res$res)
```

## Marginal-constructed barycenter

The `mcbary()` function estimates the barycenter quantile function using the
marginal-corrected barycenter procedure.

Arguments:

- `df`: the input data frame.
- `id_col`: the column identifying subjects or groups.
- `val_col`: the column containing the observed numeric values.
- `method`: the mixture estimation method; options include `"spline"`,
  `"npmle"`, `"raw"`, and `"beta"`.
- `x_grid`: the threshold grid to use when `cutpoints` is not supplied.
- `cutpoints`: the number of evenly spaced thresholds constructed from the
  observed data range; either `cutpoints` or `x_grid` must be supplied.
- `bootstrap_samples`: the number of bootstrap resamples used for uncertainty
  estimation.
- `alpha_grid`: the grid of quantile levels to estimate.
- `use_midpoint`: whether to use interval midpoints when constructing the
  quantile distribution.
- `estimate_endpoints`: whether to estimate the endpoint support regions.
- `use_isotonic_dist`: whether to enforce monotonicity on the cumulative
  mixture curves before converting them to quantile distributions.
- `use_isotonic_output`: whether to enforce monotonicity on the final
  estimated quantile curve and confidence bounds.
- `weight_col`: an optional column containing id-level weights.
- `progress`: whether to display a bootstrap progress bar.
- `ci_level`: the confidence level used for reported intervals.
- `...`: additional arguments passed to the mixture estimation routine.

```{r}
bary_res <- mcbary(
  df = sparse_dist_data,
  id_col = "id",
  val_col = "value",
  method = "spline",
  cutpoints = 20,
  bootstrap_samples = 2,
  progress = FALSE
)

head(bary_res$res)
```

The estimated quantile curve is stored in `bary_res$res`. The full object also
contains bootstrap-based uncertainty summaries, the estimated mixing
distributions, and the covariance matrix used by downstream comparison tools.

# Graphing tools

The package includes helpers for plotting both the estimated barycenter
quantiles and the intermediate mixing distributions.

```{r}
graph_quantiles(bary_res)
```

```{r}
graph_mixtures(bary_res)
```

# Comparing barycenters

To compare barycenters across samples, we first estimate each barycenter on the
same quantile grid and then apply `equalbary_test()`. The example below splits
the sample by odd and even subject identifiers only to illustrate the workflow.

```{r}
odd_ids <- unique(sparse_dist_data$id)[unique(sparse_dist_data$id) %% 2 == 1]
even_ids <- unique(sparse_dist_data$id)[unique(sparse_dist_data$id) %% 2 == 0]

sample_odd <- subset(sparse_dist_data, id %in% odd_ids)
sample_even <- subset(sparse_dist_data, id %in% even_ids)

bary_odd <- mcbary(
  sample_odd,
  id_col = "id",
  val_col = "value",
  method = "raw",
  cutpoints = 10,
  bootstrap_samples = 2,
  progress = FALSE
)

bary_even <- mcbary(
  sample_even,
  id_col = "id",
  val_col = "value",
  method = "raw",
  cutpoints = 10,
  bootstrap_samples = 2,
  progress = FALSE
)

graph_quantiles(odd = bary_odd, even = bary_even)
```

```{r}
comparison <- equalbary_test(bary_odd, bary_even, B = 50, seed = 1)

comparison$sup_norm$pval
comparison$int_sq_norm$pval
comparison$hotelling$pval
```
